---
title: "LBSPRDome: An extension of LBSPR package to include dome-shaped selectivity curves"
author: "Kristiina Hommik and Colm Fitzgerald"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 4
    number_sections: true
    toc: yes
vignette: >
  %\VignetteIndexEntry{LBSPRDome: An extension of LBSPR package to include dome-shaped selectivity curves}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Preliminaries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load vignette packages
```{r vignette_packages, warning = FALSE}
library(withr) 
library(dplyr)
library(ggplot2)
```

## Load Package
```{r}
library(LBSPRDome)
```

<!-- ## Set a random seed -->
<!-- ```{r, eval = FALSE} -->
<!-- set.seed(seed = 9999) -->
<!-- ``` -->


# GTG-LBSPR dome-shaped selectivity demonstration

We demonstrate the dome-shaped selectivity adjustment to LB-SPR methodology described in the paper
"Dome-shaped selectivity in LB-SPR: Length-Based assessment of data-limited inland fish stocks sampled with gillnets" found at <https://www.sciencedirect.com/science/article/pii/S0165783620300916> with a simple example. 

## Alternative resources
The development of the GTG-LBSPR methodology is described by Hordyk and Prince in the papers

* Hordyk *et al.* (2015) "A novel length-based empirical estimation method..." <https://doi.org/10.1093/icesjms/fsu004> and
* Hordyk *et al.* (2016) "A simple length-structured model based on life history ratios..." <https://doi.org/10.1139/cjfas-2015-0422>.

Adrian Hordyk has shared apps (e.g. <https://github.com/AdrianHordyk/LBSPR_shiny>) and code (e.g. <https://github.com/AdrianHordyk/GTG_LBSPR>) on github.



## GTG-LBSPR dome-shaped selectivity - data simulation

We first specify a set of biological stock parameters for a synthetic population

```{r StockPars, include = TRUE}
# simulated stock parameters (similar to some Irish brown trout populations) ====
StockPars <- NULL 
StockPars$MK <- 2.0
StockPars$NGTG <- 17
StockPars$Linf <- 45.0
StockPars$CVLinf <- 0.1
StockPars$MaxSD <- 2
StockPars$L50 <- 25.0  
StockPars$L95 <- 30.0 
StockPars$FecB <- 3 
StockPars$Walpha <- 0.0084
StockPars$Wbeta <- 3.1115
StockPars$Steepness <- 0.8 
StockPars$Mpow <- 0
```

and we also specify the length bins to have an increment of one unit (considered to be centimetres).

```{r SizeBins, include = TRUE}
SizeBins <- NULL
SizeBins$Linc <- 1
SizeBins$ToSize <- StockPars$Linf*(1+StockPars$MaxSD*StockPars$CVLinf)
```

The fishing gear selectivity parameters are particularly important. Instead of using a range of mesh sizes as in the paper "Dome-shaped selectivity in LBSPR...", we simply use a single unit mesh size and specify a **lognormal** dome-shaped selection curve. 

```{r FleetPars}
FleetPars <- list(selectivityCurve = "logNorm", SLmesh = 1.0)  # lognorm, single mesh size
FleetPars$SL1 <- 23.0     # k_mode from SELECT (log-normal) - mode of lognormal distribution
FleetPars$SL2 <- 0.4      # standard deviation at log scale
FleetPars$SLMin <- 0   # minimum landing limit (MLL)
```

The dome-shaped selectivity curve has a modal value of `FleetPars$SL1` and the selectivity-at-length curve is calculated as follows:
```{r calc_selectivity_at_length}

# calculate selectivity at length *IF* appropriate parameters are specified
lengthFish <- seq(0, StockPars$Linf*(1 + StockPars$CVLinf*StockPars$MaxSD), length.out = 101)
if(!is.null(FleetPars$SLmesh)) meshSize <- FleetPars$SLmesh
  
if(FleetPars$selectivityCurve=="logNorm"){ 
    SLk <- FleetPars$SL1
    SLsigma <- FleetPars$SL2
    SLMin <- FleetPars$SLMin
    
    gearSelLen <- 0
    for (j in seq_along(meshSize)){
      gearSelLen <- gearSelLen + exp(-0.5*((log(lengthFish)-log((SLk)*meshSize[j]))/(SLsigma))^2)
    }
    if(!is.na(SLMin)) gearSelLen[lengthFish < SLMin] <- 0 
    gearSelLen <- gearSelLen/max(gearSelLen)
    
}
```

to give this selectivity profile:
```{r plot_selectivity}
ggplot2::ggplot() + 
  ggplot2::geom_line(data = data.frame(length = lengthFish, selectivity = gearSelLen),
            aes(x = length, y = selectivity), colour = "black", linetype = 1, size = 1.25) + 
  ggplot2::scale_x_continuous(limits = c(0, 60), breaks = seq(0,60,10)) +
  ggplot2::theme_bw()
```


We simulate catch from this population with a prescribed fishing mortality ratio of F/M = 1.0 using multinomial random draws. The expected probabilities of a captured fish being from a particular length bin are obtained from the (per-recruit theory) expected catch-at-length for the fished/vulnerable population. This per-recruit theory is implemented in the  GTG-LBSPR simulation routine. 

```{r simulate_lbspr}
FleetPars$FM <- 1.0
simGTG <- simLBSPRDome(StockPars, FleetPars, SizeBins = NULL)
```

The length-structure of the fished and unfished catch and total population are shown below. 
```{r plot_lbspr_composition}
par(mfrow = c(2,1), mgp = c(2,1,0), mar = c(3,3,2,1) + 0.1)
# plot catch length structure
plot(simGTG$LenMids, simGTG$LCatchFished/max(simGTG$LCatchFished), 
     col = "black", type = "l", lwd = 1.5,
     xlab = "length mid", ylab = "catch proportions")
points(simGTG$LenMids, simGTG$LCatchFished/max(simGTG$LCatchFished), pch = 16, col = "black")
lines(simGTG$LenMids, simGTG$LCatchUnfished/max(simGTG$LCatchUnfished), col = "grey", type = "l", lwd = 1.5)
points(simGTG$LenMids, simGTG$LCatchUnfished/max(simGTG$LCatchUnfished), pch = 1, col = "grey")
legend("topright", c("fished", "unfished"), col = c("black", "grey"), lty = c(1,1))

# plot population length structure
plot(simGTG$LenMids, simGTG$LPopFished/max(simGTG$LPopFished), 
     col = "black", type = "l", lwd = 1.5,
     xlab = "length mid", ylab = "population (per-recruit)")
lines(simGTG$LenMids, simGTG$LPopUnfished/max(simGTG$LPopUnfished), col = "grey", type = "l", lwd = 1.5)
legend("topright", c("fished", "unfished"), col = c("black", "grey"), lty = c(1,1))

```

We then generate a random sample of `nFish` fish from a multinomial distribution with expected length bin proportions obtained from the per recruit simulation. 
```{r multinomial_sample}
nFish <- 500
# set random seed safely
withr::with_seed(999, 
                 numbersMNL <- rmultinom(1, nFish, simGTG$LCatchFished))
sampleCatch <- data.frame(lengthFish = simGTG$LenMids, 
                          numbersCatch = numbersMNL)
```

A single realisation of the multinomial draw looks like:
```{r plot_sample_catch, fig.width= 6, fig.height= 4}
ggplot2::ggplot(sampleCatch) + 
  ggplot2::geom_col(aes(x = lengthFish, y = numbersCatch), fill = "grey50") + 
  ggplot2::theme_bw()
```

This synthetic data is now used in the estimation model with both correctly specified fishing gear selectivity (dome-shaped) and mis-specified fishing gear selectivity (logistic). It is worth noting the "true" values of fishing mortality and SPR before we compare to the estimation model....
```{r}
truePars <- list(selectivityCurve = FleetPars$selectivityCurve,
                 SL1 = FleetPars$SL1, SL2 = FleetPars$SL2, SLmesh = FleetPars$SLmesh,
                 SLMin = FleetPars$SLMin, 
                 MK = StockPars$MK,
                 FM = FleetPars$FM,
                 SPR = simGTG$SPR)
truePars
```


## GTG-LBSPR dome-shaped selectivity - estimation

### True dome-shaped (lognormal) selectivity curve
We provide correctly specified biological stock parameters and (fixed) fishing gear parameters to the GTG-LBSPR estimation model.

```{r estimation_parameters}
SizeBinsEstimate <- SizeBins
StockParsEstimate <- StockPars
fixedFleetParsEstimate <- FleetPars[c("selectivityCurve", "SL1", "SL2", "SLmesh", "SLMin")]
```

The estimation model for dome-shaped selectivity curves estimates **only** the fishing mortality ratio $F/M$. The selectivity-at-length parameters are fixed.
```{r lbspr_estimation, echo = TRUE}
lbsprEstimate <- optLBSPRDome(StockPars = StockParsEstimate, 
                           fixedFleetPars = fixedFleetParsEstimate, 
                           LenDat = sampleCatch$numbersCatch, 
                           SizeBins = SizeBinsEstimate, "GTG")
```

This gives model estimates of
```{r lbspr_out}
lbsprEstimate$lbPars
```

The estimated fishing mortality ratio and SPR values are reasonably close to the true values for most realisations of the random multinomial draw of `nFish` lengths. More accurate estimation model results can be obtained by increasing the sample size of the simulated catch.

### Mis-specified selectivity curve - fixed logistic curve
The simulated length data provides a good opportunity to explore the influence of an underlying dome-shaped selectivity on LB-SPR estimates of $F/M$ and SPR when a logistic selection curve is assumed. First, we fully specify a logistic selection curve (i.e, the length-at-50% and 95% selectivity parameters are fixed and not estimated from the model). 

```{r, echo = TRUE}
FleetParsLogistic <- list(selectivityCurve = "Logistic", SL1 = 15.0, SL2 = 20.0)
SL50 <- FleetParsLogistic$SL1; 
SL95 <- FleetParsLogistic$SL2
logisticGearSelLen = 1.0/(1+exp(-log(19)*(lengthFish-SL50)/((SL95)-(SL50))))

gearSelectivityAtL <- 
  rbind(data.frame(length = lengthFish, selectivity = gearSelLen, curve = "lognorm"),
        data.frame(length = lengthFish, selectivity = logisticGearSelLen, curve = "logistic")
        )
```

The logistic selectivity parameters are chosen so that the left (ascending) limb of logistic curve closely follows that of the dome-shaped selectivity curve:
```{r compare_selectivity_curves, echo = TRUE}
ggplot2::ggplot(gearSelectivityAtL) + 
  ggplot2::geom_line(aes(x = length, y = selectivity, colour = curve), linetype = 1, size = 1.25) +
  ggplot2::scale_colour_grey(start = 0.8, end = 0.2) + 
  ggplot2::theme_bw()
```

We assess the stock status using a mis-specified logistic selection curve.
```{r lbspr_estimation_logistic, echo = TRUE}
fixedFleetParsLogistic <- FleetParsLogistic[c("selectivityCurve", "SL1", "SL2")]
# estimate FM, calculate SPR with GTG-LBSPR
lbsprLogisticEstimate <- optLBSPRDome(StockPars = StockPars, 
                           fixedFleetPars = fixedFleetParsLogistic, 
                           LenDat = sampleCatch$numbersCatch, 
                           SizeBins = SizeBins, 
                           mod = "GTG")
```

Fishing mortality and SPR are estimated as follows.
```{r}
lbsprLogisticEstimate$lbPars
```
It is evident that dome-shaped gear selection may bias estimates of fishing mortality towards higher values and estimates of SPR towards lower values when selection curves are wrongly assumed to be logistic in shape. 

### Mis-specified logistic curve - free selectivity-at-length parameters
The full estimation method for logistic selectivity, where SL50 and SL95 are free to change in the optimisation process, has not been considered yet. It can easily be handled in the dome-shaped adjustment of the method: 
```{r lbspr_full_estimation_logistic, echo = TRUE}
fixedFleetParsFullLogistic <- list(selectivityCurve = "Logistic")

lbsprFullLogisticEstimate <- optLBSPRDome(StockPars = StockPars, 
                           fixedFleetPars = fixedFleetParsFullLogistic, 
                           LenDat = sampleCatch$numbersCatch, 
                           SizeBins = SizeBinsEstimate, "GTG")
```
and gives parameters
```{r}
lbsprFullLogisticEstimate$lbPars
```

We summarise these results in a table.
```{r tabulate_data}
tableData <- rbind(
  data.frame(parameters = "true", truePars[c("selectivityCurve", "SL1", "SL2", "SLmesh", "MK", "FM", "SPR")]),
  data.frame(parameters = "estimated", fixedFleetParsEstimate[c("selectivityCurve", "SL1", "SL2", "SLmesh")], MK= StockPars$MK, FM = lbsprEstimate$lbPars[["F/M"]], SPR = lbsprEstimate$lbPars[["SPR"]]), 
  data.frame(parameters = "estimated", fixedFleetParsLogistic[c("selectivityCurve", "SL1", "SL2")], SLmesh = 1, MK= StockPars$MK, FM = lbsprLogisticEstimate$lbPars[["F/M"]], SPR = lbsprLogisticEstimate$lbPars[["SPR"]]), 
  data.frame(parameters = "estimated", fixedFleetParsFullLogistic["selectivityCurve"], SL1 = lbsprFullLogisticEstimate$lbPars[["SL50"]], SL2 = lbsprFullLogisticEstimate$lbPars[["SL95"]], SLmesh = 1, MK= StockPars$MK, FM = lbsprFullLogisticEstimate$lbPars[["F/M"]], SPR = lbsprFullLogisticEstimate$lbPars[["SPR"]])
)

colnames(tableData) <- c("parameters", "selectivityCurve", "SL1", "SL2", "SLmesh", "M/K", "F/M", "SPR")
tableData
```
